# Plotting with `ggplot2` {#sec-ggplot2}

```{r}
#| echo: false
source("_common.R")
```

__Goals:__

* Use the `ggplot2` package to make exploratory plots from STAT 113 of a single quantitative variable, a single categorical variable, two quantitative variables, a quantitative and a categorical variable, two categorical variables, and more than two variables.

* Further practice running code in `R`.

## Introduction and Basic Terminology

We will begin our data science journey with plotting in the `ggplot2` package. We are starting with plotting for a couple of reasons:

1. Plotting is cool! We get to see an immediate result of our coding efforts in the form of a nice-to-look-at plot.

1. In an exploratory data analysis, you would typically start by making plots of your data.

Throughout the first couple of sections, we will try to go very light on the technical code terminology to start out with (but we will come back to some things later in the semester). The terminology will make a lot more sense once you've actually worked with data. But, there are three terms that will be thrown around quite a bit in the next few weeks: _function_, _argument_, and _object_. 

* a _function_ in `R` is always* (*always for this class) followed by an open `(` and ended with a closed `)`. In non-technical terms, a _function_ __does__ something to its inputs and is often analogous to an English verb. For example, the `mean()` function calculates the mean, the `rank()` functions ranks a variable from lowest to highest, and the `labs()` is used to add labels to a plot. Every function has a help file that can be accessed by typing in `?name_of_function`. Try typing `?mean` in your lower left window.

* an _argument_ is something that goes inside the parentheses in a function. Arguments could include _objects_, or they might not. In the bottom-left window, type `?mean` to view the Help file on this `R` function. We see that `mean()` has 3 arguments: `x`, which is an `R` object, `trim`, and `na.rm`. `trim = 0` is the default, which means that, by default, `R` will not trim any of the numbers when computing the mean.

* an _object_ is something created in `R`, usually with `<-`. So, in the code below where we read in the STAT 113 survey data set, `stat113_df` is an `R` object. 

::: {.callout-note}
## Note

This terminology will make more sense as we go through these first couple of weeks and see some examples.
:::

## Basic Plot Structure 

We will use the `ggplot()` function in the `ggplot2` package to construct visualizations of data. the `ggplot()` function has 3 basic components:

* a `data` argument, specifying the name of your data set (`pres_df` above)
* a `mapping` argument, specifying that specifies the aesthetics of your plot (`aes()`). Common aesthetics are `x` position, `y` position, `colour`, `size`, `shape`, `group`, and `fill`.
* a `geom_    ()` component, specifying the geometric shape used to display the data.

The components are combined in the following form:

```{r}
#| eval: false

## DO NOT COPY this code into your .qmd file: it will not run.
ggplot(data = name_of_data, aes(x = name_of_x_var,
                                y = name_of_y_var,
                                colour = name_of_colour_var,
                                etc.)) +
  geom_nameofgeom() +
  .....<other stuff>
```

The structure of `ggplot()` plots is based on the Grammar of Graphics <a href="https://www.springer.com/gp/book/9780387245447" target="_blank">https://www.springer.com/gp/book/9780387245447</a>. As with most new things, the components above will be easier to think about with some examples. 

Throughout this section, we will be using survey data from STAT 113 many years ago for the exercises in this section. For those who may not have taken STAT 113 from having AP credit or another reason, the STAT 113 survey is given to all students in STAT 113 across all sections. Some analyses in Intro Stat are then carried out using the survey. 

Create a new Quarto file. Then, read in the data set by copying and pasting the `R` chunk below into an `R` chunk in your Quarto file.

```{r}
#| warning: false
library(tidyverse)
stat113_df <- read_csv("https://raw.githubusercontent.com/highamm/ds234_quarto/main/data_online/stat113.csv")
```

The data set contains the following variables:

* `Year`, FirstYear, Sophomore, Junior, or Senior
* `Sex`, M or F (for this data set, `Sex` is considered binary).
* `Hgt`, height, in inches.
* `Wgt`, weight, in pounds.
* `Haircut`, how much is paid for a haircut, typically.
* `GPA`
* `Exercise`, amount of hours of exercise in a typical week.
* `Sport`, whether or not the student plays a varsity sport.
* `TV`, amount of hours spent watching TV in a typical week.
* `Award`, Award preferred: choices are Olympic Medal, Nobel Prize, or Academy Award.
* `Pulse`, pulse rate, in beats per minute.
* `SocialMedia`, most used social media platform (Instagram, SnapChat, FaceBook, Twitter, Other, or None).

## Introduction to Plotting

We will now move into making many of the graphics that you saw in STAT 113. We will also introduce some new types of graphs that you might not have seen before. Throughout the section, we will also introduce some of the syntax used in making these plots so that, by the end, you should be able to construct an entire plot on your own.

### Single Quantitative Variable

In STAT 113, you probably used a few graphs to explore a single quantitative variable. We will begin with what is probably the most common graph to use in this situation: the __histogram__. The following code creates a histogram for the `GPA` variable in the data set named `stat113_df`.

```{r}
ggplot(data = stat113_df, aes(x = GPA)) +
  geom_histogram()
```

We can add some function __arguments__ to `geom_histogram()` to make the graph a little more pleasing to the eye. For the following histogram, we are changing the `colour`, `fill`, and the number of `bins` of the histogram of `GPA`:

```{r}
ggplot(data = stat113_df, aes(x = GPA)) +
  geom_histogram(colour = "black", fill = "white", bins = 17)
```

Let's note some of the components of the plot:

* `data = stat113_df` tells `R` which data set to use. In this case, we want to use the data set that we read in earlier and named `stat113_df`.
* `aes(x = GPA)` tells `R` to map the `GPA` variable in `stat113_df` to the `x` axis on the plot. We will come back to `aes()` (aesthetics) in a bit.
* `geom_histogram()` tells `R` that we want to use the histogram `GEOM` to bin the GPA variable and construct the histogram. The extra options `colour = "black", fill = "white", bins = 17` are options for the plot that change the `colour`, `fill`, and number of `bins` in the histogram.

There are other graphs besides a histogram that we can use to visualize a quantitative variable. Another useful graph is a frequency plot, made with the geom `geom_freqpoly()`. The code to make a frequency plot is given below.

::: {.callout-note}
## Note

The only thing that changes in the code is the `GEOM`: the data and the fact that we are mapping `GPA` to the `x`-axis remain consistent with the histogram. 
:::

```{r, appendix = TRUE}
ggplot(data = stat113_df, aes(x = GPA)) +
  geom_freqpoly(bins = 15)
```

The frequency plot is just like a histogram but the counts are connected by a line instead of represented with bins. You can see how they relate by including __both__ a `geom_freqpoly()` and a `geom_histogram()` in your plot 

::: {.callout-important}
## Important

The code below doesn't make the prettiest graph but it does show how more complex graphs are built with `ggplot()`: by continually adding __layers__ to the plot. 
:::

```{r, appendix = TRUE}
ggplot(data = stat113_df, aes(x = GPA)) +
    geom_histogram(colour = "black", fill = "white", bins = 17) +
    geom_freqpoly(bins = 15)
```

__Exercise 1__. Change the variable being plotted in the earlier histogram from `GPA` to another variable in the data set. Note that the variable must be quantitative. The variable names were given earlier in this section, but you can also see the variable names if you type in `stat113_df` in your bottom left window. Also, you can run `names(stat113_df)` in your bottom left window of `R Studio` to obtain all of the variable names in the `stat113_df` data frame.

__Exercise 2__. In the histogram you just created, change the `colour` of the histogram to be `"dodgerblue4"` and the `fill` of the histogram to be `"dodgerblue1"`.

__Exercise 3__. There are a lot of named colours that you can use in `R`. After looking through some at <a href="http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf" target="_blank">http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf</a>, change the `colour` and `fill` of the histogram again to be colours of your choice.

### Single Categorical Variable

The most common graph for visualizing a single categorical variable is a __bar plot__. The following code makes a bar plot with the `SocialMedia` variable from the STAT 113 survey. Keep in mind that the data is from 2018 - 2019, before the explosion of Tik Tok!

```{r}
ggplot(data = stat113_df, aes(x = SocialMedia)) +
  geom_bar()
```

Note the syntax used to create the plot: we still are specifying `data`, the name of the data set used (`stat113_df`) and we still are mapping a variable (`SocialMedia`) to the `x` axis of the plot. `geom_bar()` is the GEOM used to create a bar plot.

For a single categorical variable, bar plots are by far the most common plot to make so we won't discuss any other plots here.

__Exercise 4__. Choose another categorical variable to plot, and change `x = SocialMedia` to `x = name_of_your_variable`.

__Exercise 5__. `geom_bar()` also has `colour` and `fill` options just like `geom_histogram()` did. Add these arguments into `geom_bar()` in the plot you just created. You can choose whichever colours you want for `colour` and `fill`!

### Two Variables: Both Categorical 

To explore the relationship between two categorical variables, we can construct a stacked barplot. Let's look at the relationship between `Year` and `SocialMedia` first using two versions of a stacked bar plot: one stacked bar plot of the raw counts and one stacked bar plot of proportions.

To make a stacked bar plot of raw counts, we map the `SocialMedia` variable to the `fill` aesthetic:

```{r, appendix = TRUE}
ggplot(data = stat113_df, aes(x = Year, fill = SocialMedia)) +
  geom_bar() 
```

Because the number of students from each class are drastically different, it is difficult to read the distribution of `SocialMedia` preference for the levels of class Year with fewer students (Juniors and Seniors). Changing the y-axis from a count to a __proportion__ of students who prefer that particular `SocialMedia` within each class year can help mitigate this problem:

```{r}
ggplot(data = stat113_df, aes(x = Year, fill = SocialMedia)) +
  geom_bar(position = "fill") +
  labs(y = "Proportion")
```

We will not worry too much about the `position = "fill"` argument needed to make this change at this point. But, do note that, (1) with the proportion graph, we no longer see the sample size and (2) we've added `labs(y = "Proportion")` to change the y-axis label of the graph.

__Exercise 6__. Change __both__ the `x` and `fill` variables to be two other categorical variables from the STAT 113 data set. Your new plot can either be of the raw counts or of the proportion. Write a one sentence interpretation of any findings from your new plot.

### Two Variables: One Cat. and One Quant.

If we are interested in exploring the relationship between a categorical variable and a quantitative variable, side-by-side boxplots can be useful. Let's stick with the STAT 113 survey data to examine the relationship between `Exercise`, the amount of exercise in hours per week and `Award` preference.

```{r, appendix = TRUE}
ggplot(data = stat113_df, aes(x = Award, y = Exercise)) +
  geom_boxplot()
```

::: {.callout-note}
## Note

Again, examine the consistency in the syntax: we specify `data` to be the name of the data set, we map `Award` to the `x` axis, `Exercise` to the `y` axis, and we provide a relevant `GEOM`, `geom_boxplot()`.
:::

Alternatively, we can make side-by-side violin plots.

```{r, appendix = TRUE}
ggplot(data = stat113_df, aes(x = Award, y = Exercise)) +
  geom_violin()
```

The violin plots are similar to boxplots, but are wider in areas that generally have more data.

A third common option is what is known as a __strip plot__, given below:

```{r}
ggplot(data = stat113_df, aes(x = Award, y = Exercise)) +
  geom_jitter(width = 0.10)
```

The `geom_jitter()` `GEOM` adds a little bit of "noise" to the points that are plotted. In this case, the noise is added to the width and how much noise is added depends on the value of the `width` argument.

::: {.callout-important}
## Important

One big advantage of a strip plot over a side-by-side boxplots and violin plots is that we can get a rough idea about the sample size in each group in the strip plot.
:::

__Exercise 7__. Change the side-by-side violin plot to be of a different quantitative variable and a different categorical variable from the `stat113_df` data frame. Write a one sentence interpretation about any findings from your new plot.

__Exercise 8__. To your violin plot in the previous exercise, change the default `colour` and `fill` arguments to be two different colours of your choice.

### Two Quantitative Variables

For this, we will work with some fitness data collected from my Apple Watch since November 2018. The `higham_fitness_clean.csv` contains information on the following variables:

* `Date`, the month, day, and year that the fitness data was recorded on
* `month`, the month
* `weekday`, the day of the week
* `dayofyear`, the day of the year (so that 304 corresponds to the 304th day of the year)
* `distance`, distance walked in miles
* `steps`, the number of steps taken
* `flights`, the number of flights of stairs climbed
* `active_cals`, the number of calories burned from activity
* `stepgoal`, whether or not I reached 10,000 steps for the day
* `weekend_ind`, a variable for whether or not the day of the week was a weekend day (Saturday or Sunday) or a weekday (Monday - Friday).

Read in the data set by putting the following code into an `R` chunk in your Quarto file:

```{r, appendix = TRUE}
library(tidyverse)
fitness_full <- read_csv("https://raw.githubusercontent.com/highamm/ds234_quarto/main/data_online/higham_fitness_clean.csv",
                         col_types = list(stepgoal = col_factor())) |>
  mutate(weekend_ind = case_when(weekday == "Sat" | weekday == "Sun" ~ "weekend",
                                 .default = "weekday"))
```

By far, the most common plot to visualize the relationship between two quantitative variables is the scatterplot, which is made with the `geom_point()` `GEOM`. Again, in the following code, try to note the similarities in syntax with the previous plots: we still specify the `data` frame that we are using, and we still map variables in that data frame to aesthetics on the plot.

In the following scatterplot, the variable `distance` is the x-variable and `active_cals` is the y-variable.

```{r, appendix = TRUE}
ggplot(data = fitness_full, aes(x = distance, y = active_cals)) +
  geom_point()
```

__Exercise 9__. Do you see anything "odd" about the plot? Make a note of any strange observations that you see.

__Exercise 10__. Change the `x` and `y` variables plotted to two different quantitative variables in the fitness data set. Give a one sentence interpretation of any findings from your new plot. 

::: {.callout-tip collapse="true"}
## Hint

Recall from STAT 113 that three aspects you might think about are the __strength__ of the relationship (weak, moderate, strong), whether or not the relationship is __linear__, and the __direction__ of the relationship (positive or negative)
:::

### More Than Two Variables

When constructing a graph involving more than two variables, there are two commonly used strategies: (1) Use another plot aesthetic for the third variable (most commonly a __colour__ aesthetic) and (2) Facet by the third variable.

First, in addition to `x` and `y` aesthetics, we can also use `aes()` to map variables to things like `colour`, `size`, and `shape`. For example, using the STAT 113 data set again, we might make a scatterplot with `Wgt` on the x-axis and `Hgt` on the y-axis, and use the `colour` aesthetic for the `Sex` of the student. In this older version of the STAT 113 survey there are only two levels of `Sex` in the data: `M` and `F`.

```{r, appendix = TRUE}
ggplot(data = stat113_df, aes(x = Wgt, y = Hgt, colour = Sex)) +
  geom_point()
```

Colour is the most popular aesthetic, but other aesthetics include `shape` (if the third variable is categorical) and `size` (if the third variable is quantitative). Neither of these are particularly useful for this example but examples of using each are given in the next code chunk.

```{r}
ggplot(data = stat113_df, aes(x = Wgt, y = Hgt, shape = Sex)) +
  geom_point()
ggplot(data = stat113_df, aes(x = Wgt, y = Hgt, size = Exercise)) +
  geom_point()
```

A second option for including a third variable in a plot is to __facet__ by that variable using the `facet_wrap()` function. The syntax of `facet_wrap()` is `facet_wrap(~ name_of_faceting_variable)`. 

::: {.callout-important}
## Important

Faceting creates a __subplot__ for each __level__ of the variable that you choose to facet by.
:::

For example, in the following plot, we make a scatterplot of `Hgt` vs. `Wgt` and facet by `Year`. This makes 4 subplots: one for each class `Year`:

```{r}
ggplot(data = stat113_df, aes(x = Wgt, y = Hgt)) +
  geom_point() +
  facet_wrap(~ Year)
```

In general, the more levels of the variable there are, the more useful faceting is compared to using a `colour` aesthetic. We will explore this idea more in the Exercises.

__Exercise 11__. In the `fitness_full` data, construct a scatterplot of two of the quantitative variables and facet by one of the categorical variables.

## Practice

In general, practice exercises will be split between exercises that will be done together as a class and exercises that you will do in groups or on your own. The purpose of the class exercises is to give some guidance on how we might think logically through some of the code and the results. The purpose of the group exercises and the on your own exercises is so that you have a chance to practice what you've learned with your table or on your own.

### Class Exercises

__Class Exercise 1__. As we progress through the semester, we will pay close attention to the way that our data set is structured. In the earlier bar plot that we made of the `SocialMedia` variable from the `stat113_df` data frame, we had a column called `SocialMedia` where each row represented a student. `geom_bar()` then counts up the total number of times each level appears to make the bar plot. 

However, sometimes, data are in format such that one column contains the _levels_ of the categorical variable while another column contains the counts directly. For example, we can create such a data set using code that we will learn next week and name it `stat113_restructured`:

```{r}
stat113_restructured <- stat113_df |> group_by(SocialMedia) |>
  summarise(n_social = n())
stat113_restructured
```

a. Suppose that you attempted to make a bar plot of `SocialMedia` with `stat113_restructured` using `geom_bar()`. Predict what the plot will look like.

b. Construct the plot described in (a) to test your prediction.

c. When the data is structured such that there is a column that uniquely gives the levels of the categorical variable and a second column that gives the count of each level, we typically use the `geom_col()` `GEOM` instead of `geom_bar()`. `geom_col()` uses both an `x` and a `y` aesthetic mapping. Construct the bar plot using `stat113_restructured` with the `geom_col()` `GEOM`.

__Class Exercise 2__*. What is the relationship between distance and active calories?

<!-- After adding the smoother, investigate the following arguments: (1) `method`, (2) `se`, and (3) `span`. -->


__Class Exercise 3__*. What is the relationship between number of steps taken and date? Is there an increasing trend or a decreasing trend or neither? 

__Class Exercise 4__*. What is the relationship between GPA and Exercise? Is that relationship different depending on sex? depending on sport? depending on sex-sport combination?

__Class Exercise 5__. In STAT 113, boxplots are typically introduced using the `*` symbol to identify outliers. Using a combination of the help `?geom_boxplot` and Googling "R point shapes", figure out how to modify your side-by-side boxplots so that the outliers are shown using `*`, not the default dots. 

__Class Exercise 6__. We can change the y-axis of a histogram to be "density" instead of a raw count. This means that each bar shows a __proportion__ of cases instead of a raw count. Google something like "ggplot2 geom_histogram with density" to figure out how to create a y `aes()` to show density instead of count.

### Your Turn

__Your Turn 1__. Suppose that you ask each student in this class the following question: Is your major `Statistics`, `Data Science`, or `Other`. You now want to make a bar plot of your hypothetical data. In pairs, have one person complete part (a) and the second person complete part (b).

a. Sketch out (with pen and paper) how the data would be structured if you were to use `geom_bar()` to make a bar plot of the `Major` variable.

b. Sketch out (with pen and paper) how the data would be structured if you were to use `geom_col()` to make a bar plot of the `Major` variable.

__Your Turn 2__. With a partner, decide on __two__ quantitative variables and __one__ categorical variable from the STAT 113 data set that you are interested in exploring. Have one person complete part (a) and the second person complete part (b).

a. Make a scatterplot that colours the points by the categorical variable.

b. Make a set of faceted scatterplots (that are faceted by the levels of the categorical variable). 

With your partner, decide which of the two graphs is "better" for exploring the three variables and why.

__Your Turn 3__. Repeat the previous exercise with your partner with different variables, but switch parts (so that you will work on (a) if you previously worked on (b) and vice versa).

__Your Turn 4__. Besides strip plots, boxplots, and violin plots, we can also make coloured frequency plots or faceted histograms if we are interested in exploring the relationship between a quantitative variable and a categorical variable. With a partner, decide on one quantitative and one categorical variable that you wish to explore from the `fitness_full` data set. Have one person complete part (a) and the second person complete part (b).

a. Make a coloured frequency plot of the variables you selected (colouring by the levels of the categorical variable).

b. Make a set of faceted histograms (faceting by the categorical variable that you selected).

c. Do you prefer either of these graphs to the "standard" side-by-side boxplots for the variables that you selected?

__Your Turn 5__. In your group, discuss why it would not make sense to construct a line plot of `Exercise` vs. `GPA` from the `stat113_df` data set.

::: {.callout-tip collapse="true"}
## Hint

What about the `GPA` variable in particular would make construction of a line plot impossible?
:::

__Your Turn 6__. In your group, discuss why you think `R` gives us a warning message that it "Removed 70 rows containing non-finite values" whenever we make a plot of the `GPA` variable from `stat113_df`. 

__Your Turn 7__. On your own, construct a plot of your choosing, with any variables from either of the data sets we have used so far! Give a one sentence interpretation of your findings from your plot.

__Your Turn 8__. A common theme that we'll see throughout the course is that it's advantageous to know as much background information as possible about the data set we are analyzing. Data sets will be easier to analyze and pose questions about if you're familiar with the subject matter. 

In your group, give an example of something that you know about STAT 113 and the survey data set that helped you answer or pose a question that someone from another university (and therefore unfamiliar with our intro stat course) wouldn't know.

In your group, give an example of something that you don't know about the fitness data set that the person who owns the fitness data would know. Why does that give an advantage to the person who is more familiar with the fitness data?

## Aesthetic Mapping in `ggplot`

__Goals__:

* further use `ggplot2` to construct plots of data
* explain how aesthetics are used in `ggplot()` and describe the difference between local and global aesthetics.

The purpose of this second section on `ggplot2` is to discuss more about plot aesthetics. While the previous section focused more on basic plot construction, this section will focus more on how the plots are constructed by what is specified in the `aes()` mapping argument.

::: {.callout-important}
## Important

The aesthetic mappings are what map __variables__ in the data frame we are using to characteristics of the plot.
:::

Which aesthetic mappings are available depend on the `GEOM` used. Common mappings include `x`, `y`, `colour`, `fill`, `size`, `shape`, and `linewidth`.

However, `colour`, `fill`, `size`, `shape`, and `linewidth` are also things that we commonly change outside of aesthetic mappings. Details of when to put an argument inside of `aes()` or outside of `aes()` are discussed next.

### Inside vs. Outside `aes()`

Examine the following graph of some of the variables in the fitness data set we used in the previous section.

```{r}
library(tidyverse)
fitness_full <- read_csv("https://raw.githubusercontent.com/highamm/ds234_quarto/main/data_online/higham_fitness_clean.csv",
                         col_types = list(stepgoal = col_factor())) |>
  mutate(weekend_ind = case_when(weekday == "Sat" | weekday == "Sun" ~ "weekend",
                                 .default = "weekday")) |>
  mutate(season = case_when(dayofyear >= 355 | dayofyear <= 80 ~ "winter",
                            dayofyear >= 81 & dayofyear <= 172 ~ "spring",
                            dayofyear >= 173 & dayofyear <= 264 ~ "summer",
                            .default = "fall"))

ggplot(data = fitness_full,
       aes(x = Date, y = steps, colour = season)) +
  geom_point()
```

Here, we have used a `colour` aesthetic to map the `season` variable to `colour` so that we can more easily distinguish points from the various seasons in the plot. But, what if we wanted to simply colour all of the points `"blue"`. Try running:

```{r}
ggplot(data = fitness_full,
       aes(x = Date, y = steps, colour = "blue")) +
  geom_point()
```

What does the graph look like? Did it do what you expected? To actual colour the points `"blue"`, we need to specify `colour = "blue"` __outside__ of `aes()` because `"blue"` is not a variable from the `fitness_full` data frame. Also, when giving non-aesthetic characteristics to a plot, the argument must go with the associated `GEOM`; so, `colour = "blue"` must be an argument to `geom_point()`:

```{r}
ggplot(data = fitness_full,
       aes(x = Date, y = steps)) +
  geom_point(colour = "blue")
```

To sum up, putting `colour = ____` inside `aes()` or outside `aes()` achieves different things. In general,

* when we want to map something in our data set (`fitness_full`) to something in our plot (`x`, `y`, `colour`, `size`, etc.), we put that __inside__ the `aes()` as in `aes(colour = weekend_ind)`. 

* When we assign fixed characteristics (making all points a larger size, widening the width of a line, using a constant colour, etc.) that don't come from the data, we put them __outside__ of `aes()`.

::: {.callout-note}
## Note

We actually have already used non-aesthetic options in `ggplot()`, though we did not frame them as non-aesthetic options at the time. In the previous section, we often changed the `colour` option and `fill` option to histograms, bar plots, and box plots. These are non-aesthetic options because, again, we are not mapping a variable in the data frame to these arguments.
:::

__Exercise 1__. The default point `size` is `1.5` while the default point `shape` is `19`. In the following plot, change the defaults so that the points are a little larger than the default and so that the shape is an `X`. Google image search "`R point shapes`" to figure out which integer corresponds to the `X` point shape.

```{r}
ggplot(data = fitness_full,
       aes(x = Date, y = steps)) +
  geom_point(size = 1.5, shape = 19)
```

__Exercise 2__. Use the `linewidth` option in `geom_smooth()` to increase the width of the smoother. Note that the default `linewidth` is `1`.

```{r}
ggplot(data = fitness_full,
       aes(x = Date, y = steps)) +
  geom_point() +
  geom_smooth()
```

### Local vs. Global Aesthetics

Up until this point, we have always supplied aesthetics __globally__ in the `ggplot()` function. However, there are times when we must override the global aesthetics and change the __locally__ within an individual `GEOM`. 

In the following plot, the __global__ aesthetics `x`, `y`, and `colour` are passed to all further `GEOM`s (`geom_point()` and `geom_smooth()` in this case) unless specifically overridden. So, the plot with global aesthetics

```{r}
ggplot(data = fitness_full,
       aes(x = Date, y = steps, colour = season)) +
  geom_point() +
  geom_smooth()
```

is equivalent to the following plot with local aesthetics:

```{r}
ggplot(data = fitness_full) +
  geom_point(aes(x = Date, y = active_cals, colour = season)) +
  geom_smooth(aes(x = Date, y = active_cals, colour = season))
```

::: {.callout-important}
## Important

When we put the `aes()` in `ggplot()` as global aesthetics, `R` perpetuates these `aes()` aesthetics in all further `GEOM`s in your plotting command. 
:::

As our plots become more complex, we will need to think carefully about whether we want to supply an aesthetic globally or locally. As an example, we can change which aesthetics are supplied globally and which are supplied locally to create a graph that has the points coloured by `season` but a single smoother and a graph that has smoothers for each level of `season` but a single point colour:

```{r}
ggplot(data = fitness_full,
       aes(x = Date, y = steps)) +
  geom_point(aes(colour = season)) +
  geom_smooth()

ggplot(data = fitness_full,
       aes(x = Date, y = steps)) +
  geom_point() +
  geom_smooth(aes(colour = season))
```

Which graph you choose depends on what you might want to emphasize in your plot!

__Exercise 3__. Modify the following plot so that the aesthetics are all explicitly given __locally__ to each `GEOM` instead of __globally__.

```{r}
ggplot(data = fitness_full,
       aes(x = Date, y = steps)) +
  geom_point() +
  geom_line()
```

### `R` Code Style

We want our code to be as readable as possible. This not only benefits other people who may read your code (like me), but it also benefits you, particularly if you read your own code in the future. I try to follow the Style Guide in the Advanced `R` book: <a href="http://adv-r.had.co.nz/Style.html" target="_blank">http://adv-r.had.co.nz/Style.html</a>. Feel free to skim through that, but you don't need to worry about it too much: you should be able to pick up on some important elements just from going through this course. You might actually end up having better code style if you _haven't_ had any previous coding experience.

As a quick example of why code style can be important, consider the following two code chunks, both of which produce the same graph.

```{r, eval = FALSE}
ggplot(data=fitness_full,aes(x=Date,y=steps))+geom_point(colour=
     "darkgreen")+geom_smooth(se=FALSE,colour="black",linewidth=1.4)
```

```{r, eval = FALSE}
ggplot(data = fitness_full,
       aes(x = Date, y = steps)) +
  geom_point(colour = "darkgreen") +
  geom_smooth(se = FALSE, colour = "black", linewidth = 1.4)
```

Which code chunk would you want to read two years from now? Which code chunk would you want your classmate/friend/coworker to read? (assuming you like your classmate/friend/coworker....)

__Exercise 4__. Fix the following code so that the code is more readable.

```{r}
ggplot(data =fitness_full,
       aes(x =weekday,y= steps)) +geom_boxplot(fill="coral1",colour ="black")
```

__Exercise 5__. Read the very short paper at <a href="https://joss.theoj.org/papers/10.21105/joss.01686" target="_blank">https://joss.theoj.org/papers/10.21105/joss.01686</a> on an Introduction to the `tidyverse`, and answer the questions below in your `Quarto` file. Answer the following questions by typing answers in your .qmd document. Because these answers do not involve code, you should be able to type them into your `.qmd` document outside of a code chunk.

a. What are the two major areas that the `tidyverse` __doesn't__ provide tools for?

b. How do the authors define "tidy"?

c. What does it mean for the `tidyverse` to be "human-centred"?

d. In about 2 sentences, describe the data science "cycle" given in the diagram at the top of page 3. 

## Practice

In general, practice exercises will be split between exercises that will be done together as a class and exercises that you will do in groups or on your own. The purpose of the class exercises is to give some guidance on how we might think logically through some of the code and the results. The purpose of the group exercises and the on your own exercises is so that you have a chance to practice what you've learned with your table or on your own.

For some of these exercises we will again use the `stat113_df` data, described in the previous section.

Moreover, we will start to use the `glimpse()` function more often to get a very quick look at our data:

```{r}
stat113_df <- read_csv("https://raw.githubusercontent.com/highamm/ds234_quarto/main/data_online/stat113.csv")
glimpse(stat113_df)
```

What does `glimpse()` tell us?

### Class Exercises

__Class Exercise 1__. With the `stat113_df` data, make a scatterplot of `Hgt` on the y-axis and `Wgt` on the x-axis, colouring by `Sport`. Add a smooth fitted curve to your scatterplot. Then, move `colour = Sport` from an `aes()` in the `ggplot()` function to an `aes()` in the `geom_point()` function. What changes in the plot? Can you give an explanation as to why that change occurs?

__Class Exercise 2__. Fix the following plot so that the fill colour of the histogram is `"white"` and the outline colour is `"black"`.

```{r}
ggplot(data = stat113_df, aes(x = TV)) +
  geom_histogram(aes(colour = "black", fill = "white"))
```

### Your Turn

__Your Turn 1__. Suppose you have a generic data frame named `toy_df`, with quantitative variables `xvar` and `yvar`, and categorical variable `groupvar`. In your group, select which of the following plots will result in a scatterplot of `xvar` and `yvar` where there are separate coloured fitted linear regression line for each level of `groupvar` but all of the points are coloured `"purple"`. You may select one, more than one, or none at all.

```{r}
#| eval: false
## a
ggplot(data = toy_df, aes(x = xvar, y = yvar, colour = groupvar)) +
  geom_point(colour = "purple") +
  geom_smooth(aes(method = "lm"))

## b
ggplot(data = toy_df, aes(x = xvar)) +
  geom_point(aes(y = yvar), colour = "purple") +
  geom_smooth(aes(y = yvar, colour = groupvar), method = "lm")

## c
ggplot(data = toy_df, aes(x = xvar, y = yvar)) +
  geom_point(aes(colour = "purple")) +
  geom_smooth(method = "lm", colour = groupvar)

## d
ggplot(data = toy_df, aes(x = xvar, y = yvar)) +
  geom_point(colour = "purple") +
  geom_smooth(method = "lm", colour = groupvar)

## e
ggplot(data = toy_df, aes(x = xvar, y = yvar)) +
  geom_point(colour = "purple") +
  geom_smooth(method = "lm", aes(colour = groupvar))

## f
ggplot(data = toy_df, aes(x = xvar, y = yvar)) +
  geom_point(aes(colour = "purple")) +
  geom_smooth(method = "lm", aes(colour = groupvar))
```

__Your Turn 2__. Choose one of the plots that you selected as "correct" in the previous exercise. In your group, discuss, how the syntax of the plot would change if all aesthetics were supplied __locally__ instead of __globally__? Based on this discussion, when are global aesthetics more useful than local aesthetics?
