# Coding in Base `R`

__Goals__:

* describe common classes for variables in a data set.
* explain why some `R` errors come about from class misspecifications.
* use indexing to reference rows, columns, or specific observations in a `tibble` or data set.

__Motivation__

"Base `R`" generally refers to `R` code that we can use without loading in any outside packages (so this code is __not__ in the `tidyverse` family of packages). Why is the chapter on `R` basics not the first chapter that we discuss? There certainly are advantages of doing things that way, but there are also advantages of not starting out with something like "classes of variables in `R`." 

First, it's not the most inherently interesting thing to look at. It's a lot more fun to make plots and wrangle data. As long as someone makes sure that the variables are already of the "correct" class, then there's no need to talk about this.

Second, much of what we discuss here will make more sense, having the previous four chapters under our belt. We'll be able to see how misspecified variable classes cause issues in certain summaries and plots _and we already know how to make those plots and get those summaries_.

## Variable Classes in `R`

`R` has a few different classes that variables take, including `numeric`, `factor`, `character` `Date`, and `logical`. Before we delve into the specifics of what these classes mean, let's try to make some plots to illustrate why we should care about what these classes mean.

The `videogame_clean.csv` file contains variables on video games from 2004 - 2019, including

* `game`, the name of the game
* `release_date`, the release date of the game
* `release_date2`, a second coding of release date
* `price`, the price in dollars,
* `owners`, the number of owners (given in a range)
* `median_playtime`, the median playtime of the game
* `metascore`, the score from the website Metacritic
* `price_cat`, 1 for Low (less than 10.00 dollars), 2 for Moderate (between 10 and 29.99 dollars), and 3 for High (30.00 or more dollars)
* `meta_cat`, Metacritic's review system, with the following categories: "Overwhelming Dislike", "Generally Unfavorable", "Mixed Reviews", "Generally Favorable", "Universal Acclaim".
* `playtime_miss`, whether median play time is missing (`TRUE`) or not (`FALSE`)

The data set was modified from <a href="https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-07-30" target="_blank">https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-07-30</a>.

Run the code in the following `R` chunk to read in the data.

```{r, message = FALSE, appendix = TRUE}
library(tidyverse)
library(here)
videogame_df <- read_csv(here("data/videogame_clean.csv"))
head(videogame_df)
```

A data frame or tibble holds variables that are allowed to be different classes. If a variable is a different class than you would expect, you'll get some strange errors or results when trying to wrangle the data or make graphics. 

Run the following lines of code. In some cases, we are only using the first 100 observations in `videogame_small`. Otherwise, the code would take a __very__ long time to run.

```{r, appendix = TRUE}
videogame_small <- videogame_df |> slice(1:100)
ggplot(data = videogame_small, aes(x = release_date, y = price)) +
  geom_point() 

ggplot(data = videogame_small, aes(x = release_date2, y = metascore)) +
  geom_point(aes(colour = price_cat))
```

In the first plot, `release_date` isn't ordered according to how you would expect (by date). Instead, `R` orders it alphabetically.

In the second plot, we would expect to get a plot with 3 different colours, one for each level of `price_cat`. Instead, we get a continuous colour scale, which doesn't make sense, given that `price_cat` can only be `1`, `2`, or `3`.

Both plots are not rendered correctly because the variable classes are not correct in the underlying data set. Up until this point, the data that has been provided has almost always had the correct variable classes, by default, but that won't always be the case!

We've actually seen both of these issues before as well (the Date issue in the exercise data and the continuous colour scale in the cars data), but, in both of these instances, code was provided to "fix" the problem. After this section, you'll have the tools to fix many `class` issues on your own!

If you examine the output of the following line of code

```{r, appendix = TRUE}
head(videogame_df)
```

you'll see that, at the very top of the output, right below the variable names, `R` provides you with the classes of variables in the `tibble`. 

* `<chr>` is character, used for strings or text.
* `<fct>` is used for variables that are factors, typically used for character variables with a finite number of possible values the variable can take on.
* `<date>` is used for dates.
* `<dbl>` stands for double and is used for the `numeric` class.
* `<int>` is for numbers that are all integers. In practice, there is not much difference between this class and class `dbl`.
* `<lgl>` is for logical, variables that are either `TRUE` or `FALSE`.

### Referencing Variables and Using `str()`

We can use `name_of_dataset$name_of_variable` to look at a specific variable in a data set:

```{r, results = "hide", appendix = TRUE}
videogame_df$game
```

prints the first thousand entries of the variable `game`. There are a few ways to get the `class` of this variable: the way that we will use most often is with `str()`, which stands for "structure", and gives the class of the variable, the number of observations (26688), as well as the first couple of observations:

```{r, appendix = TRUE}
str(videogame_df$game)
```

You can also get a variable's class more directly with `class()`

```{r, appendix = TRUE}
class(videogame_df$game)
```

## Classes in Detail

The following gives summary information about each class of variables in `R`:

### `<chr>` and `<fct>` Class

With the character class, `R` will give you a warning and/or a missing value if you try to perform any numerical operations:

```{r, appendix = TRUE}
mean(videogame_df$game)
videogame_df |> summarise(maxgame = max(game))
```

You also can't convert a character class to numeric. You can, however, convert a character class to a `<fct>` class, using `as.factor()`. The `<fct>` class will be useful when we discuss the `forcats` package, but isn't particularly useful now.

```{r, appendix = TRUE}
class(videogame_df$meta_cat)
class(as.factor(videogame_df$meta_cat))
```

In general, `as._____` will lets you convert between classes. Note, however, that we aren't saving our converted variable anywhere. If we wanted the conversion to the factor to be saved in the data set, we can use `mutate()`:

```{r, appendix = TRUE}
videogame_df <- videogame_df |>
  mutate(meta_cat_factor = as.factor(meta_cat))
str(videogame_df$meta_cat_factor)
```

For most `R` functions, it won't matter whether your variable is in class character or class factor. In general, though, character classes are for variables that have a ton of different levels, like the name of the videogame, whereas factors are reserved for categorical variables with a finite number of levels.

### `<date>` Class

The `<date>` class is used for dates, and the `<datetime>` class is used for Dates with times. `R` requires a very specific format for dates and times. Note that, while to the human eye, both of the following variables contain dates, only one is of class `<date>`:

```{r, appendix = TRUE}
str(videogame_df$release_date)
str(videogame_df$release_date2)
```

`release_date` is class character, which is why we had the issue with the odd ordering of the dates earlier. You can try converting it using `as.Date`, but this function doesn't always work:

```{r, error = TRUE}
as.Date(videogame_df$release_date)
```

Dates and times can be pretty complicated. In fact, we will spend an entire week covering them using the `lubridate` package.

On variables that are in `Date` format, like `release_date2`, we can use numerical operations:

```{r, appendix = TRUE}
median(videogame_df$release_date2, na.rm = TRUE)
mean(videogame_df$release_date2, na.rm = TRUE)
```

What do you think taking the median or taking the mean of a date class means?

### `<dbl>` and `<int>` Class

Class `<dbl>` and `<int>` are probably the most self-explanatory classes. `<dbl>`, the `numeric` class, are just variables that have only numbers in them while `<int>` only have integers (..., -2, -1, 0, 1, 2, ....). You can do numerical operations on either of these classes (and we've been doing them throughout the semester). For our purposes, these two classes are interchangeable.

```{r, appendix = TRUE}
str(videogame_df$price)
```

Problems arise when numeric variables are coded as something non-numeric, or when non-numeric variables are coded as numeric. For example, examine:

```{r, appendix = TRUE}
str(videogame_df$price_cat)
```

`price_cat` is categorical but is coded as `1` for cheap games, `2` for moderately priced games, and `3` for expensive games. Therefore, `R` thinks that the variable is numeric, when, it's actually a factor.

```{r, appendix = TRUE}
str(as.factor(videogame_df$price_cat))
```

This is the cause of the odd colour scale that we encountered earlier and can be fixed by converting `price_cat` to a factor:

```{r, appendix = TRUE}
videogame_df <- videogame_df |>
  mutate(price_factor = as.factor(price_cat)) 
ggplot(data = videogame_df, aes(x = release_date2, y = metascore)) +
  geom_point(aes(colour = price_factor))
```

### `<lgl>` Class

Finally, there is a class of variables called logical. These variables can only take 2 values: `TRUE` or `FALSE`. For example, `playtime_miss`, a variable for whether or not the `median_playtime` variable is missing or not, is logical:

```{r, appendix = TRUE}
str(videogame_df$playtime_miss)
```

It's a little strange at first, but `R` can perform numeric operations on logical classes. All `R` does is treat every `TRUE` as a `1` and every `FALSE` as a `0`. Therefore, `sum()` gives the total number of `TRUE`s and `mean()` gives the proportion of `TRUE`s. So, we can find the number and proportion of games that are missing their `median_playtime` as:

```{r, appendix = TRUE}
sum(videogame_df$playtime_miss)
mean(videogame_df$playtime_miss)
```

There's a lot of games that are missing this information! 

We've actually used the ideas of logical variables for quite some time now, particularly in statements involving `if_else()`, `case_when()`, `filter()`, and `mutate()`.

The primary purpose of this section is to be able to identify variable classes and be able to convert between the different variable types with `mutate()` to "fix" variables with the incorrect class.

### Exercises {#exercise-6-2}

Exercises marked with an \* indicate that the exercise has a solution at the end of the chapter at \@ref(solutions-6).

We will use the fitness data set again for this set of exercises, as the data set has some of the issues with variable class that we have discussed. However, in week 1, some of the work of the work to fix those issues was already done before you saw the data. Now, you'll get to fix a couple of those issues! Read in the data set with:

```{r, message = FALSE}
library(tidyverse)
fitness_df <- read_csv(here("data/higham_fitness_notclean.csv"))
```

1. \* What is the issue with the following plot? After you figure out the issue, use `mutate()` to create a new variable that fixes the issue and then reconstruct the graph.

```{r, echo = TRUE}
ggplot(data = fitness_df, aes(x = active_cals)) +
  geom_freqpoly(aes(group = weekday, colour = weekday))
```

2. \* What is another variable in the data set that has an incorrect `class`?

3. Create a new variable, called `step_goal` that is `1` or `TRUE` if at least 10000 steps were walked and `0` or `FALSE` if fewer than 10000 steps were walked. Using this variable, find the total number of days where the goal was met and the proportion of the days where the goal was met.

## Object Types and Subsetting

Variables of these different classes can be stored in a variety of different objects in `R`. We have almost exclusively used the `tibble` object type. The tidy `tibble`

* is "rectangular" and has a specific number of rows and columns.
* has columns that are variables
* each column must have elements that are of the same class, but different columns can be of different classes. This allows us to have character and numeric variables in the same `tibble`.

### `tibble` and `data.frame`

The `tibble` object is very similar to the `data.frame` object. You can also check what type of object you're working with using the `str()` command:

```{r, results = "hide", appendix = TRUE}
str(videogame_df) ## look at the beginning to see "tibble"
```

We will have a small section on `tibble`s in the coming weeks so we won't focus on them here. But, we should take note that, to reference a specific element in a `tibble`, called _indexing_, you can use `[# , #]`. So, for example, `videogame_df[5, 3]` grabs the value in the fifth row and the third column:

```{r, appendix = TRUE}
videogame_df[5, 3]
```

More often, we'd want to grab an entire row (or range of rows) or an entire column. We can do this by leaving the row number blank (to grab the entire column) or by leaving the column number blank (to grab the entire row):

```{r, results = "hide", appendix = TRUE}
videogame_df[ ,3] ## grab the third column

videogame_df[5, ] ## grab the fifth row
```

We can also grab a range of columns or rows using the `:` operator:

```{r, results = "hide", appendix = TRUE}
3:7

videogame_df[ ,3:7] ## grab columns 3 through 7

videogame_df[3:7, ] ## grab rows 3 through 7
```

or we can grab different columns or rows using `c()`:

```{r, results = "hide", appendix = TRUE}
videogame_df[ ,c(1, 3, 4)] ## grab columns 1, 3, and 4

videogame_df[c(1, 3, 4), ] ## grab rows 1, 3, and 4
```

To get rid of an entire row or column, use `-`: `videogame_df[ ,-c(1, 2)]` drops the first and second columns while `videogame_df[-c(1, 2), ]` drops the first and second rows.

### Vectors

A vector is an object that holds "things", or elements, of the same class. You can create a vector in `R` using the `c()` function, which stands for "concatenate". We've used the `c()` function before to bind things together; we just hadn't yet discussed it in the context of creating a vector.

```{r, appendix = TRUE}
vec1 <- c(1, 3, 2)
vec2 <- c("b", 1, 2)
vec3 <- c(FALSE, FALSE, TRUE)
str(vec1); str(vec2); str(vec3)
```

Notice that `vec2` is a character class. `R` requires all elements in a vector to be of one class; since `R` knows `b` can't be numeric, it makes all of the numbers characters as well.

Using the `dataset$variable` draws out a vector from a `tibble` or `data.frame`:

```{r, results = "hide", appendix = TRUE}
videogame_df$metascore
```

If you wanted to make the above vector "by hand", you'd need to have a lot of patience: `c(96, 88, 65, NA, NA, NA, 93, .........)`

Just like tibbles, you can save vectors as something for later use:

```{r, appendix = TRUE}
metavec <- videogame_df$metascore
mean(metavec, na.rm = TRUE)
```

How would you get the mean `metascore` using `dplyr` functions?

Vectors are one-dimensional: if we want to grab the 100th element of a vector we just use `name_of_vector[100]`:

```{r, appendix = TRUE}
metavec[100] ## 100th element is missing
```

Be aware that, if you're coming from a math perspective, a "vector" in `R` doesn't correspond to a "vector" in mathematics or physics.

### Lists

Lists are one of the more flexible objects in `R`: you can put objects of different classes in the same list and lists aren't required to be rectangular (like `tibbles` are). Lists are __extremely__ useful because of this flexibility, but, we won't use them much in this class. Therefore, we will just see an example of a list before moving on:

```{r, appendix = TRUE}
testlist <- list("a", 4, c(1, 4, 2, 6),
                 tibble(x = c(1, 2), y = c(3, 2)))
testlist
```

`testlist` has four elements: a single character `"a"`, a single number `4`, a vector of `1, 4, 2, 6`, and a `tibble` with a couple of variables. Lists can therefore be used to store complex information that wouldn't be as easily stored in a `vector` or `tibble`.

### Exercises {#exercise-6-3}

Exercises marked with an \* indicate that the exercise has a solution at the end of the chapter at \@ref(solutions-6).

1. \* Look at the subsetting commands with `[ , ]`. What `dplyr` functions can you use to do the same thing?

2. Create a `tibble` called `last100` that only has the last 100 days in the data set using both (1) indexing with `[ , ]` and (2) a `dplyr` function.

3. Create a tibble that doesn't have the `flights` variable using both (1) indexing with `[ , ]` and (2) a `dplyr` function. 

4. \* Use the following steps to create a new variable `weekend_ind`, which will be "weekend" if the day of the week is Saturday or Sunday and "weekday" if the day of the week is any other day. The current `weekday` variable is coded so that `1` represents Sunday, `2` represents Monday, ...., and `7` represents Saturday. 

a. Create a vector that has the numbers corresponding to the two weekend days. Name the vector and then create a second vector that has the numbers corresponding to the five weekday days.

b. Use `dplyr` functions and the `%in%` operator to create the new `weekend_ind` variable. You can use the following code chunk to help with what `%in%` does:

```{r, echo = TRUE, results = "hide"}
1 %in% c(1, 2, 3, 4)
2 %in% c(1, 2, 3, 4)

2 %in% c(3, 4, 5, 6)
```

## Other Useful Base `R` Functions

In addition to functions like `%in%` in the previous exercise, there are many other useful base `R` functions. The following give some of the functions that I think are most useful for data science. 

__Generating Data__: `rnorm()`, `sample()`, and `set.seed()`

`rnorm()` can be used to generate a certain number of normal random variables with a given mean and standard deviation. It has three arguments: the sample size, the mean, and the standard deviation.

`sample()` can be used to obtain a sample from a vector, either with or without replacement: it has two required arguments: the vector that we want to sample from and `size`, the size of the sample.

`set.seed()` can be used to fix `R`'s random seed. This can be set so that, for example, each person in our class can get the same random sample as long we all set the same seed.

These can be combined to quickly generate toy data. For example, below we are generating two quantitative variables (that are normally distributed) and two categorical variables:

```{r, appendix = TRUE}
set.seed(15125141)
toy_df <- tibble(xvar = rnorm(100, 3, 4),
                 yvar = rnorm(100, -5, 10),
                 group1 = sample(c("A", "B", "C"), size = 100, replace = TRUE),
                 group2 = sample(c("Place1", "Place2", "Place3"), size = 100,
                                 replace = TRUE))
toy_df
```

__Tables__: We can use the `table()` function with the `$` operator to quickly generate tables of categorical variables:

```{r, appendix = TRUE}
table(toy_df$group1)

table(toy_df$group1, toy_df$group2)
```

__Others__: There are quite a few other useful base `R` functions. `nrow()` can be used on a data frame to quickly look at the number of rows of the data frame and `summary()` can be used to get a quick summary of a vector:
  
```{r, appendix = TRUE}
nrow(toy_df)
summary(toy_df$yvar)
```

There are also some useful functions for viewing a data frame. `View()` function can be used in your console window on a data frame: `View(toy_df)` to pull up a spreadsheet-like view of the data set in a different window within `R Studio`.

Options to `print()` allow us to view more rows or more columns in the console printout:

```{r, appendix = TRUE}
toy_df |>
  print(n = 60) ## print out 60 rows
toy_df |>
  print(width = Inf) ## print out all of the columns
```

We will stop here, but will surely encounter more base `R` functions as we run into different types of problems.

### Exercises {#exercise-6-4}

Exercises marked with an \* indicate that the exercise has a solution at the end of the chapter at \@ref(solutions-6).

1. Use `dplyr` and `tidyr` functions to re-create the tables generated from

```{r}
table(toy_df$group1)

table(toy_df$group1, toy_df$group2)
```

## Chapter Exercises {#chapexercise-6}

Exercises marked with an \* indicate that the exercise has a solution at the end of the chapter at \@ref(solutions-6).

Work through the following exercises pertaining to the video game data set.

1. \* Read in the data set and use `filter()` to remove any rows with missing metascores, missing median playtime, or have a median playtime of 0 hours.

__Note__: We usually don't want to remove missing values without a valid reason. In this case, a missing metascore means that the game wasn't "major" enough to get enough critic reviews, and a missing or 0 hour median playtime means that there weren't enough users who uploaded their playtime to the database. Therefore, any further analyses are constructed to games that are popular enough to both get enough reviews on metacritic and have enough users upload their median playtimes.

```{r, message = FALSE}
videogame_df <- read_csv(here("data/videogame_clean.csv"))
```

2. \* Make a scatterplot with `median_playtime` on the y-axis and `metascore` on the x-axis with the filtered data set. 

3. \* Something you may notice is that many of the points directly overlap one another. This is common when at least one of the variables on a scatterplot is _discrete_: `metascore` can only take on integer values in this case. Change `geom_point()` in your previous plot to `geom_jitter()`. Then, use the help to write a sentence about what `geom_jitter()` does.

4. \* Another option is to control point transparency with `alpha`. In your `geom_jitter()` statement, change `alpha` so that you can still see all of the points, but so that you can tell in the plot where a lot of points are overlapping.

5. \* Label the points that have median playtimes above 1500 hours. You may want to use the `ggrepel` package so that the labels don't overlap.

6. Choose one of the games that got labeled and Google that game's median, or possibly average, play time. Is it in the vicinity as the `median_playtime` recorded in our data set? 

```{r, echo = FALSE}
## for every game, I Googled, the answer is "no". The median or average
## play time is much less than what's given in our data set
```

7. What should be done about the outliers? We will discuss and investigate this issue as a class.

```{r, echo = FALSE}
## main points: what's a reasonable median playtime? You don't want to throw away a point that's not a data entry error: that point could be interesting! But you don't want to keep a point that's obviously an error. Transparency is __key__. 
```

## Exercise Solutions {#solutions-6}

### Variable Classes S

### Classes in Detail S

1. \* What is the issue with the following plot? After you figure out the issue, use `mutate()` to create a new variable that fixes the issue and then reconstruct the graph.

```{r, fig.keep = "last"}
ggplot(data = fitness_df, aes(x = active_cals)) +
  geom_freqpoly(aes(group = weekday, colour = weekday))
```

The issue is that weekday should be a factor, not numeric.

```{r}
fitness_df <- fitness_df |> mutate(weekday_cat = as.factor(weekday))
ggplot(data = fitness_df, aes(x = active_cals)) +
  geom_freqpoly(aes(group = weekday_cat, colour = weekday_cat)) +
  scale_colour_viridis_d()
```

2. \* What is another variable in the data set that has an incorrect `class`?

Month should be an ordered factor, not numeric.

### Object Types S

1. \* Look at the subsetting commands with `[ , ]`. What `dplyr` functions can you use to do the same thing?

`slice()` can be used for the row indexing while `select()` can be used for the column indexing.

4. \* Use the following steps to create a new variable `weekend_ind`, which will be "weekend" if the day of the week is Saturday or Sunday and "weekday" if the day of the week is any other day. The current `weekday` variable is coded so that `1` represents Sunday, `2` represents Monday, ...., and `7` represents Saturday. 

a. Create a vector that has the numbers corresponding to the two weekend days. Name the vector and then create a second vector that has the numbers corresponding to the five weekday days.

```{r}
vecweekend <- c(1, 7)
vecweekday <- 2:6 ## or c(2, 3, 4, 5, 6)
```

b. Use `dplyr` functions and the `%in%` operator to create the new `weekend_ind` variable. You can use the following code chunk to help with what `%in%` does:

```{r, echo = TRUE}
1 %in% c(1, 2, 3, 4)
2 %in% c(1, 2, 3, 4)

2 %in% c(3, 4, 5, 6)
```

```{r}
fitness_df |>
  mutate(weekend_ind = case_when(weekday %in% vecweekend ~ "weekend",
                                 weekday %in% vecweekday ~ "weekday")) |>
  select(weekend_ind, everything())

## can also use if_else, which is actually a little simpler in this case:
fitness_df |> mutate(weekend_ind = if_else(weekday %in% vecweekend,
  true = "weekend", false = "weekday")) |>
  select(weekend_ind, everything())
```

### Chapter Exercises S {#chapexercise-6-S}

1. \* Read in the data set and use `filter()` to remove any rows with missing metascores, missing median playtime, or have a median playtime of 0 hours.

__Note__: We usually don't want to remove missing values without a valid reason. In this case, a missing metascore means that the game wasn't "major" enough to get enough critic reviews, and a missing or 0 hour median playtime means that there weren't enough users who uploaded their playtime to the database. Therefore, any further analyses are constructed to games that are popular enough to both get enough reviews on metacritic and have enough users upload their median playtimes.

```{r, message = FALSE}
videogame_df <- read_csv(here("data/videogame_clean.csv"))
```

```{r}
videogame_nomiss <- videogame_df |>
  filter(!is.na(median_playtime) &
           !is.na(metascore) &
           median_playtime != 0)
```

2. \* Make a scatterplot with `median_playtime` on the y-axis and `metascore` on the x-axis with the filtered data set. 

```{r}
ggplot(data = videogame_nomiss, aes(x = metascore,
                                    y = median_playtime)) + 
  geom_point()
```

3. \* Something you may notice is that many of the points directly overlap one another. This is common when at least one of the variables on a scatterplot is _discrete_: `metascore` can only take on integer values in this case. Change `geom_point()` in your previous plot to `geom_jitter()`. Then, use the help to write a sentence about what `geom_jitter()` does.

```{r}
ggplot(data = videogame_nomiss, aes(x = metascore,
                                    y = median_playtime)) + 
  geom_jitter()
```

`geom_jitter()` adds a small amount of "noise" to each data point so that points don't overlap quite as much.

4. \* Another option is to control point transparency with `alpha`. In your `geom_jitter()` statement, change `alpha` so that you can still see all of the points, but so that you can tell in the plot where a lot of points are overlapping.

```{r}
ggplot(data = videogame_nomiss, aes(x = metascore,
                                    y = median_playtime)) + 
  geom_jitter(alpha = 0.4)
## can see a lot of ponits have median playtimes close to 0
```

5. \* Label the points that have median playtimes above 1500 hours. You may want to use the `ggrepel` package so that the labels don't overlap.

```{r}
library(ggrepel)
videogame_long <- videogame_nomiss |> filter(median_playtime > 1500)
ggplot(data = videogame_nomiss,
       aes(x = metascore, y = median_playtime)) + 
  geom_jitter(alpha = 0.4) +
  geom_label_repel(data = videogame_long, aes(label = game))
```

## Non-Exercise `R` Code {#rcode-6}

```{r ref.label=knitr::all_labels(appendix == TRUE), echo=TRUE, eval=FALSE}
```
