# Factors with `forcats` {#forcats}

```{r}
#| echo: false
source("_common.R")
```

__Goals:__

* Use the `forcats` package to change the levels of factors, or to re-order levels of factors in a way that makes tables and graphs easier to read.

## Change Factor Levels

__The Data__: The `pokemon_allgen.csv` data set contains observations on Pokemon from the first 6 Generations (the first 6 games). There are 20 variable in this data set, but, of particular interest for this chapter are

* `Type 1`, the first Type characteristic of the Pokemon (a factor with 13 levels)
* `Type 2`, the second Type characteristic of the Pokemon (a factor with 13 levels, `NA` if the Pokemon only has one type)
* `Generation`, the generation the Pokemon first appeared in (a factor with 6 levels)

Read in the data set with `read_csv()`. Then, use a `mutate()` statement to make a `Generation_cat` variable that is a factor.

```{r, message = FALSE, warning = FALSE, appendix = TRUE}
library(tidyverse)
library(here)
pokemon_df <- read_csv(here("data/pokemon_allgen.csv")) |>
  mutate(Generation_cat = factor(Generation))
```

One easy way to get a quick summary of a `factor` variable is to use `group_by()` and `n()` within a `summarise()` statement:

```{r, appendix = TRUE}
pokemon_df |> group_by(`Type 1`) |>
  summarise(counttype = n())
```

<br>

### `fct_recode()` to Rename Levels

Now, let's make a bar plot that examines how many `Legendary` Pokemon first appear in each generation, using `dplyr` commands that we've used and a simple `geom_col()` `GEOM`:

```{r, appendix = TRUE}
pokemon_legend <- pokemon_df |> filter(Legendary == TRUE) |>
  group_by(Generation_cat) |>
  summarise(nlegend = n())
ggplot(data = pokemon_legend, aes(x = Generation_cat, y = nlegend)) +
  geom_col() +
  theme_minimal()
```

We've discussed how to change many aspects of `ggplot2` graphs, but we haven't discussed how to rename the labels of levels of a categorical variable, whether those appear in the x-axis or in a separate legend. The easiest way to do this is to rename the levels in the factor itself using `fct_recode()`. Suppose, for example, that we want to relabel the Generation number with the actual region corresponding to each game (`Kanto`, `Johto`, `Hoenn`, `Sinnoh`, `Unova`, and `Kalos`). The function `fct_recode()` takes the name of a factor already present in the data set as its first argument and then a series of renaming schemes (new_name = "old_name") as its remaining arguments.

```{r, appendix = TRUE}
pokemon_legend <- pokemon_legend |>
  mutate(Generation_cat2 = fct_recode(Generation_cat,
                                      Kanto = "1", Johto = "2",
                                      Hoenn = "3", Sinnoh = "4",
                                      Unova = "5", Kalos = "6")) |>
  select(Generation_cat2, everything())
pokemon_legend

ggplot(data = pokemon_legend,
       aes(x = Generation_cat2, y = nlegend)) +
  geom_col() +
  theme_minimal()
```

### Collapsing Many Levels Into Fewer Levels with `fct_collapse()`

Sometimes, we might want to collapse the levels of two or more factors into a single level. With the Pokemon data set, there isn't an example where this really makes sense, but, in the exercises, we'll see a good use for this function with the social survey data set. For practice, we can collapse the `Ice` and `Dark` type Pokemon into a new level called `Coolest` and we can collapse the `Poison`, `Fighting`, and `Fire` type Pokemon into a new level called `Least_Cool`. 

```{r, appendix = TRUE}
pokemon_long <- pokemon_df |> pivot_longer(c(`Type 1`, `Type 2`),
                            names_to = "Number",
                            values_to = "Type")
pokemon_long |>
  mutate(new_type = fct_collapse(Type, Coolest = c("Ice", "Dark"),
                                 Least_Cool = c("Fire", "Fighting", "Poison"))) |>
  select(new_type, Type, everything())
```

What happens to the levels that aren't being re-specified?

<br>

__Exercise 1__. What `dplyr` function(s) could we also use to create the new levels that were created with `fct_collapse()`? Why might it be a little easier to use `fct_collapse()`? 

__Exercise 2__. Use `fct_recode()` with `mutate()` to recode "A" to "Apple", and "B" to "Banana" in the following toy data set:

```{r}
fruit_df <- tibble::tibble(fruit = c("A", "B", "B", "A"))
```

## Reorder Factor Levels

### Change the Order of Levels by a Quantitative Variable with `fct_reorder()`

We might also be interested in re-ordering the x or y-axis of a particular graph so that the order of the factors correspond to, for example, the median of a quantitative variable for each level. The reason we would want to do this is easiest to see with an example. For example, suppose we want to look at the most common Pokemon types across the first 6 generations. We first get rid of duplicate Pokemon (we'll discuss this in an exercise). Then, we pivot the data so that type is in one column, and we remove observations with missing `Type`, which correspond to the second `Type` of Pokemon that only have a single `Type`:

```{r, appendix = TRUE}
pokemon_nodup <- pokemon_df |> group_by(`#`) |> slice(1) |>
  ungroup()
pokemon_long <- pokemon_nodup |>
  pivot_longer(c(`Type 1`, `Type 2`),
               names_to = "Number",
               values_to = "Type")
pokemon_sum <- pokemon_long |>
  group_by(Type) |>
  summarise(count_type = n()) |>
  filter(!is.na(Type))
ggplot(data = pokemon_sum, aes(x = Type,
                               y = count_type)) +
  geom_col() +
  coord_flip() + ## flips the x and y axes
  theme_minimal()
```

How does `R` order the levels of the `Type` factor, by default? How might you like them to be ordered to make the graph more readable?

The following code creates a new factor variable called `Type_ordered` that orders type by the `count_type` variable. `fct_reorder()` takes a factor as its first argument and a numeric variable to re-order that factor by as its second argument. The bar plot is then reconstructed with this new variable.

```{r, appendix = TRUE}
pokemon_sum <- pokemon_sum |> 
  mutate(Type_ordered = fct_reorder(.f = Type, .x = count_type))
ggplot(data = pokemon_sum, aes(x = Type_ordered,
                               y = count_type)) +
  geom_col() +
  coord_flip() +
  theme_minimal()
```

### Lollipop Plots

Lollipop plots are a popular alternative to bar plots because they often look cleaner with less ink. To make a lollipop plot in `R`, we specify two different geoms: `geom_segment()` to form the stick of the lollipop and `geom_point()` to form the pop part of the lollipop. `geom_segment()` requires 4 aesthetics: `x`, `xend`, `y`, and `yend`.

```{r, appendix = TRUE}
ggplot(data = pokemon_sum, aes(x = Type_ordered,
                               y = count_type)) +
  geom_segment(aes(x = Type_ordered, xend = Type_ordered,
                   y = 0, yend = count_type)) +
  geom_point() +
  coord_flip() +
  theme_minimal()
```

`fct_reorder()` also works with boxplots or simple point plots that show, for example, the median response for each level of a factor. The following set of plots investigate how the `Defense` stat changes for different Pokemon types

```{r, appendix = TRUE}
pokemon_long <- pokemon_long |>
  filter(!is.na(Type)) |>
  mutate(Type_Deford = fct_reorder(.f = Type, .x = Defense,
                                   .fun = median))
ggplot(data = pokemon_long, aes(x = Type_Deford,
                               y = Defense)) +
  geom_boxplot() + 
  coord_flip() +
  theme_minimal()
```

The following code makes a point plot that shows the median defense for each type instead of boxplots.

```{r, appendix = TRUE}
pokemon_med <- pokemon_long |> group_by(Type_Deford) |>
  summarise(med_def = median(Defense)) |>
  mutate(Type_Deford = fct_reorder(.f = Type_Deford, .x = med_def,
                                   .fun = median))

ggplot(data = pokemon_med, aes(x = med_def, y = Type_Deford)) +
  geom_point() +
  theme_minimal()
```

Finally, we can make a lollipop plot of median defense.

```{r, appendix = TRUE}
ggplot(data = pokemon_med, aes(x = Type_Deford, y = med_def)) +
  geom_segment(aes(xend = Type_Deford, y = 0, yend = med_def)) +
  geom_point() +
  coord_flip() +
  theme_minimal()
```

::: {.callout-note}
## Note

The previous two plots (the point plot of median defense and the lollipop plot of median defense) are really not the best types of plots to explore defense across the different Pokemon types. We will discuss why as part of a class exercise.
:::

__New Data__. The `gun_violence_us.csv` data set was obtained from <a href="https://www.openintro.org/book/statdata/index.php?data=gun_violence_us" target="_blank">https://www.openintro.org/book/statdata/index.php?data=gun_violence_us</a> and contains the following variables on gun violence in 2014:

* `state`, the name of the U.S. state
* `mortality_rate`, number of deaths from gun violence per 100,000 people
* `ownership_rate`, the proportion of adults who own a gun
* `region`, region of the U.S. (`South`, `West`, `NE`, and `MW`)

```{r, message = FALSE, appendix = TRUE}
mortality_df <- read_csv(here("data/gun_violence_us.csv")) |>
  mutate(region = factor(region))
```

### Reordering Levels Manually with `fct_relevel()`

Suppose that we want to investigate the relationship between `mortality_rate` and `ownership_rate` using this data set. Run the following code to create a scatterplot of `mortality_rate` vs. `ownership_rate` with fitted linear regression lines for each `region` of the United States:

```{r, message = FALSE, appendix = TRUE}
ggplot(data = mortality_df,
       aes(x = ownership_rate, y = mortality_rate, colour = region)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```

Notice the order of the levels in the legend. Most people would prefer the order to actually match up with where the lines in the plot end, not for the order to be alphabetical. To achieve this, we can reorder the levels manually with `fct_relevel()`.

Factors are ordered alphabetically by default. If we want precise control over the order of the levels of a factor, we can use `fct_relevel()`, which takes a factor and a vector of the new levels as inputs:

```{r, message = FALSE, appendix = TRUE}
mortality_df <- mortality_df |>
  mutate(region_3 = fct_relevel(region, c("South", "West", "MW", "NE")))
ggplot(data = mortality_df,
       aes(x = ownership_rate, y = mortality_rate, colour = region_3)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```

Reordering the levels of a factor manually might also be useful in fitting linear models. Recall that, by default, `R` makes the __reference group__ in a linear model the first level alphabetically. If we'd like a different reference group, you can reorder the levels of the factor:

```{r, results = "hide", appendix = TRUE}
mod <- lm(mortality_rate ~ ownership_rate + region, data = mortality_df)
mod3 <- lm(mortality_rate ~ ownership_rate + region_3, data = mortality_df)
summary(mod)
summary(mod3)
```

If you have not taken STAT 213, you can ignore this discussion about the reference group in linear models. 

__Exercise 3__. The `.fun` argument in `fct_reorder()` controls how the `Type` factor is ordered. In the boxplots of Pokemon defense, change this argument when making `pokemon_long` so that the boxplots are ordered by the maximum `Defense` instead of the median `Defense`.

```{r}
ggplot(data = pokemon_long, aes(x = Type_Deford,
                               y = Defense)) +
  geom_boxplot() + 
  coord_flip() +
  theme_minimal()
```

## Practice

### Class Exercises

__Class Exercise 1__. In the text, it is mentioned that the point plot of median defense and the lollipop plot of median defense are not great visualizations. Why? For what type of data would these visualizations be more appropriate? 

__Class Exercise 2__. For the pokemon data set, we mentioned that we got rid of duplicates with `group_by(`#`) |> slice(1) |> ungroup()`. Explain why the code would get rid of duplicates. We will then discuss a more convenient `dplyr` function we can use here.

__Class Exercise 3__. Make the side-by-side boxplots again with the pokemon data but do not use `ungroup()` by running the following code.

```{r}
pokemon_nodup <- pokemon_df |> group_by(`#`) |> slice(1) ## |>
  ## ungroup()
pokemon_long <- pokemon_nodup |>
  pivot_longer(c(`Type 1`, `Type 2`),
               names_to = "Number",
               values_to = "Type")

pokemon_long <- pokemon_long |>
  filter(!is.na(Type)) |>
  mutate(Type_Deford = fct_reorder(.f = Type, .x = Defense,
                                   .fun = median))
ggplot(data = pokemon_long, aes(x = Type_Deford,
                               y = Defense)) +
  geom_boxplot() + 
  coord_flip()
```

Why are the boxplots no longer ordered by median defense?

__Class Exercise 4__. On a feedback sheet, you were asked to "rank" this course in terms of challenge and amount of time you spend on it, compared to the other courses you are taking this semester. This is a question that has been asked of students in past sections of this course. Read in some data collected in a couple of past semesters:

```{r}
responses_df <- read_csv("https://raw.githubusercontent.com/highamm/ds234_quarto/main/data_online/heatmap.csv")
responses_df
```

<!-- Our goal with this data set is to visualize the distribution of responses to this question with a __heatmap__. To do so, we will: -->

<!-- 1. Figure out how to construct a heatmap using `geom_tile()`: <https://r-graph-gallery.com/79-levelplot-with-ggplot2.html> -->

<!-- 2. Change the x and y-axis labels of the heatmap to be `"Most Challenging"`, `"Second"`, `"Third"`, and `"Least Challenging"` and `"Most Time"`, etc. -->

<!-- 3. Make sure the ordering of the levels in the x and y axes make sense. -->

<!-- 4. Make the plot look a little "prettier" by using some other `ggplot2` options we have not yet discussed in detail. -->

### Your Turn

For these your turn exercises, we will use a data set on National Football League standings from 2000 to 2020. Read in the data set with:

```{r}
library(tidyverse)
library(here)
standings_df <- read_csv(here("data/standings.csv"))
standings_df
```

The important variables that we will use include:

* `team`, the city where the team is based in
* `team_name`, the name of the team
* `playoffs`, whether or not the team made the playoffs that year
* `sb_winner`, whether or not the team won the superbowl that year

__Your Turn 1__. Use the `table()` function with `table(name_of_data_frame$name_of_variable)` to make a table of `team_name`. This is useful to use for categorical variables to give a quick summary of what the levels are and how many times each level appears in the data set.

__Your Turn 2__. Until a few years ago, the Washington Commanders team used to be known as the Washington Redskins. Because of the obvious racism the name conveys, in 2022, the name was changed from Redskins to Commanders. Use a `forcats` function to rename the `Redskins` `team_name` to `Commanders`. Note that, usually, we have been renaming the new variable after we use a `forcats` function, but, oftentimes, it makes sense to just overwrite the old variable by using the same name in our `mutate()` statement.

__Your Turn 3__. Use a function from `tidyr` to combine `team` and `team_name` into a single variable called `franchise`. You may want to specify `sep = " "` for consistency with the city names.

__Your Turn 4__. There are a couple of franchises in the national football league that moved cities in the late 2010s. In particular, the San Diego Chargers became the Los Angeles Chargers and the St. Louis Rams became the Los Angeles Rams (this is another instance where being familiar with context is helpful here: it may have taken you much longer to figure this out, had you not known much about the NFL). Use a `forcats` function to put the `San Diego Chargers` and `Los Angeles Chargers` into a single level, `San Diego LA Chargers`, and to put the `St. Louis Rams` and `Los Angeles Rams` into a single level, `St. Louis LA Rams`.

__Your Turn 5__. Using the updated data set, make a plot that helps visualize which franchises have made the playoffs most often during this time span. Feel free to customize your plot with <https://r-graph-gallery.com/301-custom-lollipop-chart.html>.

```{r, echo = FALSE, eval = FALSE}
standings_df <- standings_df |> mutate(team_name = fct_recode(team_name,
                                              Commanders = "Redskins")) |>
  unite(col = "franchise", c(team, team_name), sep = " ",
        remove = FALSE) |>
  mutate(franchise = fct_collapse(franchise,
                                  `St. Louis LA Rams` =  c("St. Louis Rams", "Los Angeles Rams"),
                                  `San Diego LA Chargers` = c("Los Angeles Chargers", "San Diego Chargers")))

top_ten <- standings_df |> filter(playoffs == "Playoffs") |>
  group_by(franchise) |>
  summarise(n_playoffs = n()) |>
  arrange(desc(n_playoffs)) |>
  slice(1:10) |>
  mutate(franchise = fct_reorder(franchise, n_playoffs))
ggplot(data = top_ten, aes(x = franchise, y = n_playoffs)) +
  geom_point() +
  geom_segment(aes(x = franchise, xend = franchise,
                   y = 0, yend = n_playoffs)) +
  coord_flip()
```

__Your Turn 6__. Which is more strongly associated with a team's win percentage: offensive ranking or defensive ranking? (We will not dive into how these offensive and defensive metrics are calculated exactly, but, in general, a larger ranking means that the team's offense (or defense) is better).

```{r}
#| echo: false
#| output: false
ggplot(data = standings_df, aes(x = offensive_ranking, y = wins / (wins + loss))) +
  geom_point()
ggplot(data = standings_df, aes(x = defensive_ranking, y = wins / (wins + loss))) +
  geom_point()
cor(standings_df$offensive_ranking, standings_df$wins)
cor(standings_df$defensive_ranking, standings_df$wins)
```


The following are additional exercises for `forcats`: I've left them in here in case you want some extra practice!
We will use the general social survey data set, which is in the `forcats` library in `R`. You should some of this Wikipedia page to better understand where this data comes from <a href="https://en.wikipedia.org/wiki/General_Social_Survey" target="_blank">Wikipedia</a>.

Most variables are self-explanatory, but a couple that aren't are:

* `partyid`, political leaning and 
* `denom`, religious denomination (if unfamiliar with this, you can think of it as a "more specific" subset of a particular religion).

Note that some of these exercises are from the R for Data Science textbook.

Load in the data set with

```{r, echo = TRUE, results = "hide"}
library(tidyverse)
gss_cat
```

__Additional Exercise 1__. Using a `forcats` function, change the name of the level `Not str republican` to be `Weak republican` and change the name of the level `Not str democrat` to be `Weak democrat`. These names more closely match the levels `Strong republican` and `Strong democrat`. Then, create a table of counts that shows the number of respondents in each political party `partyid`. 

__Note__: Levels that aren't specified in your `forcats` function do not change.

__Note 2__: In naming something `Weak republican`, you'll need to use backticks since there is a space in the level name.

__Additional Exercise 2__. Use a `forcats` function so that `partyid` just has 4 categories: Other (corresponding to No answer, Don't know, Other party), Ind (corresponding to Ind,near rep, Independent, Ind, near dem), Rep (corresponding to Strong republican and Not str republican), and Dem (corresponding to Not str democrat and Strong democrat).

__Additional Exercise 3__. Run the code to create the following plot that shows the average number of hours of television people watch from various religions. 

```{r, fig.keep = "none"}
relig_summary <- gss_cat |>
  group_by(relig) |>
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(data = relig_summary, aes(tvhours, relig)) +
  geom_point()
```

Then, use a `forcats` function create a new variable in the data set that reorders the religion factor levels and make a lollipop plot so that the religion watches the most television, on average, is on the top, and the religion that watches the least television, on average, is on the bottom.

__Additional Exercise 4__. Run the code to make the following line plot that shows age on the x-axis, the proportion on the y-axis, and is coloured by various marital statuses (married, divorced, widowed, etc.):

```{r, fig.keep = "none"}
by_age <- gss_cat |>
  filter(!is.na(age)) |>
  count(age, marital) |>
  group_by(age) |>
  mutate(prop = n / sum(n))

ggplot(by_age, aes(age, prop,
                  colour = marital)) +
  geom_line(na.rm = TRUE) +
  labs(colour = "marital")
```

Then, use a `forcats` function to make the plot so that the legend labels line up better with the different coloured marital status lines (e.g. so that the label for widowed is the first that appears in the legend, the label for married is second, etc.).

__Additional Exercise 5__. We haven't talked much about creating two-way tables (or contingency tables). These are generally quite difficult to make with the `tidyverse` functions, but you can use the base `R` `table()` and `prop.table()` functions to make these.

Using data only from the year 2014, run the following code to make 4 two-way tables with the `party_small` variable that was constructed earlier and `race`:

```{r}
gss_cat <- gss_cat |> mutate(party_small = fct_collapse(partyid,
                                              Other = c("No answer", "Don't know", "Other party"),
                                              Ind = c("Ind,near rep", "Independent", "Ind,near dem"),
                                              Rep = c("Strong republican", "Not str republican"),
                                              Dem = c("Not str democrat", "Strong democrat")))

gss_recent <- gss_cat |> filter(year == 2014)

tab1 <- table(gss_recent$party_small, gss_recent$race)
tab1
prop.table(tab1)
prop.table(tab1, margin = 1)
prop.table(tab1, margin = 2)
```

Use the help on `?prop.table` to figure out how each of these three tables are constructed.

Which table do you think is most informative? What conclusions does it help you to draw?

