# Predictive Modeling with knn {#knn}

```{r}
#| echo: false
source("_common.R")
```

__Goals__

* explain why it's necessary to use _training data_ and _test data_ when building a predictive model.
* describe the k-nearest neighbors (knn) procedure.
* interpret a _classification table_.
* use knn to predict the levels of a categorical response variable for _test data_.

::: {.callout-note}
## Note

The structure of this section will be a bit different than the structure of the previous sections. We will complete some of this material on handouts that we will fill in by-hand.
:::

## Introduction to Classification

We will introduce both the knn algorithm and classification in general using a handwritten handout.

## Choosing Predictors and k

We will continue to use the scaled version of the pokemon data set for this handout. This time, we will have __75__ pokemon in our training data set and we are only looking at the `Steel`, `Dark`, `Fire`, and `Ice` types. As discussed in the handout, it is important to scale all of our numeric predictors so that the unit of measurement does not influence the classification results. We can scale all of the numeric variables in a data frame using a combination of `across()`, `where()`, and `mutate()` before we split the data into a training sample and a test sample.

```{r}
set.seed(1119)
library(tidyverse)
library(pander)
library(here) 

pokemon <- read_csv(here("data", "pokemon_full.csv")) |>
  filter(Type %in% c("Steel", "Dark", "Fire", "Ice")) |>
  mutate(across(where(is.numeric), ~ (.x - min(.x)) /
                                 (max(.x) - min(.x)))) 
```

Next, we split the data into a traning sample of 75 pokemon and a test sample.

```{r}
train_sample <- pokemon |>
  slice_sample(n = 75)

test_sample <- anti_join(pokemon, train_sample)
```

There are many candidate predictors in this data set: `HP`, `Attack`, `Defense`, ..., all the way up to `base_experience`. How should we determine which predictors to include in our model? 

Much of this will be trial and error by evaluating different models with a criterion that we will talk about in the next section. However, it is always helpful to explore the data set with graphics to get us to a good starting point. A __scatterplot matrix__ is a useful exploratory tool. The following is a scatterplot matrix with the response variable, `Type`, and just three candidate predictors, `HP`, `Attack`, and `Defense`, created with the `GGally` ("g-g-ally") package.

```{r}
## install.packages("GGally")
library(GGally)
ggpairs(data = train_sample, columns = c(4, 5, 6, 3))
```

The `columns` argument is important: it allows you to specify which columns you want to look at. I prefer putting the response, `Type` (column `3`) in the last slot.

We can examine this to see which variables seem to have a relationship with `Type`. Where would we want to look for this?

What's given on the diagonal of the scatterplot matrix?

Which variables might we want to include as predictors in a knn model? 

__Exercise 1__. Construct another scatterplot matrix with `Type` and a different set of predictors. Which predictors look like they might be useful to include in a knn model to predict `Type`?

After we decide on an initial set of predictors to include, we'll use the `class` package to fit a knn model in `R`. For our first model, let's just use `HP`, `Attack` and `Defense` as predictors. The `kknn` library can fit knn models with a `kknn()` function using syntax similar to `lm()`, which is used to fit linear models in STAT 213. 

```{r, appendix = TRUE}
## install.packages("kknn")
library(kknn)
```

The arguments to `kknn()` are 

* `formula`, which gives a formula in the style `response ~ pred1 + pred2 + .... + predx`. For example, if we wanted `Attack`, `Defense` and `Speed` to be predictors for `Type`, our formula would be `Type ~ Attack + Defense + Speed`.
* `train`, a data set with the training data.
* `test`, a data set with the test data.
* `k`, the number of nearest neighbors we would like to use in our model.

For those that have taken or are enrolled in STAT 213, this is quite similar to the way you fit regression models with a couple of differences: the function requires __2__ data sets instead of __1__ (as it requires both a training and a test data set) and the function also requires that we specify a value for $k$, the number of nearest neighbors to use in our model. The example below fits a knn model with `Attack`, `Defense`, and `Speed` as predictors using $k = 9$ nearest neighbors.

```{r, appendix = TRUE}
#| echo: true
#| eval: false
## fit the knn model with 9 nearest neighbors
knn_mod <- kknn(Type ~ Attack + Defense + Speed, train = train_sample,
                test = test_sample, k = 9)
knn_mod
```

If you try to run the code above, you will actually get an error! This is one of the rare examples in this class that we see a difference in the way an `R` function treats `<chr>` variables and `<fct>` variables. For the `kknn()` function to run, the response variable must be a __factor__ and is not allowed to be a __character__. Let's make the conversion and then refit the model. Note that, because we have to make the conversion in both the training and the test sample, it will be easier to convert the variable type to factor before we do the split into training and test:

```{r}
set.seed(1119)

pokemon <- read_csv(here("data", "pokemon_full.csv")) |>
  filter(Type %in% c("Steel", "Dark", "Fire", "Ice")) |>
  mutate(across(where(is.numeric), ~ (.x - min(.x)) /
                                 (max(.x) - min(.x)))) |>
  mutate(Type = as.factor(Type))

train_sample <- pokemon |>
  slice_sample(n = 75)

test_sample <- anti_join(pokemon, train_sample)

knn_mod <- kknn(Type ~ Attack + Defense + Speed, train = train_sample,
                test = test_sample, k = 9)
knn_mod
```

The output of `knn_mod` just gives the formula that we used to fit the data and tells us that the response is "nominal" (which just means that the response is categorical). If we want to examine what the actual classifications are for Pokemon in the test sample, we can pull these out with:

```{r}
knn_mod$fitted.values
```

So, the first pokemon in the test sample is predicted to be `Fire` type, the second is predicted to be `Fire` type, etc, the fifth is predicted to be `Ice` type, etc.

## Evaluating a Predictive Model

But, how well did our model classify pokemon into `Type`s? We still need a metric to evaluate models with different predictors. One definition of a "good" model in the classification context is a model that has a high proportion of correct predictions in the test data set. This should make some intuitive sense, as we would hope that a "good" model correctly classifies most `Dark` pokemon as `Dark`, most `Fire` pokemon as `Fire`, etc. 

In order to examine the performance of a particular model, we'll create a __classification table__ that shows the results of the model's classification on observations in the test data set. An equivalent name for the confusion matrix is a __confusion matrix__.

We can compare the predictions from the knn model with the actual pokemon `Type`s in the test sample with `table()`, which makes the classification table:

```{r, appendix = TRUE}
table(knn_mod$fitted.values, test_sample$Type) 
```

The columns of the classification table give the actual Pokemon types in the test data while the rows give the predicted types from our knn model. 

__Exercise 2__. Interpret the value of `12` in the classification table above.

__Exercise 3__. Interpret the value of `3` in the column with `Dark` and the row with `Ice`.

__Exercise 4__. Interpret the value of `0` in the bottom-left of the classification table above.

One common metric used to assess overall model performance is the model's __classification rate__, which is computed as the number of correct classifications divided by the total number of observations in the test data set. 

__Exercise 5__. Compute the classification rate "by hand" (that is, by using `R` as a calculator).

Code to automatically obtain the classification rate from a confusion matrix is

```{r}
#| output: false
tab <- table(knn_mod$fitted.values, test_sample$Type) 
sum(diag(tab)) / sum(tab)
```

What does `diag()` seem to do in the code above?

__Exercise 6__. Change the predictors used or change k to improve the classification rate of the model with `k = 9` and `Attack`, `Defense`, `HP`, and `Speed` as predictors.

__Exercise 7__. A baseline classification rate to compare to is a model that just classifies everything in the test data set as the most common `Type` in the training data set. In this case, what would the "baseline" classification rate be?

```{r}
#| echo: false
#| output: false

train_sample |> group_by(Type) |>
  summarise(n = n())
## baseline classification would classify every Pokemon in the test sample 
## as Fire

test_sample |>
  mutate(correct_fire = if_else(Type == "Fire",
                                true = 1,
                                false = 0)) |>
  relocate(correct_fire) |>
  summarise(prop_fire = mean(correct_fire))
```

__Exercise 8__. We will choose $k$, the number of neighbors considered, using a bit of trial and error, but we will also automate the process by writing a for loop to loop through different values of $k$. However, we should discuss the relative advantages of smaller and larger k values. Which value is "best" is entirely dependent on the data at hand! What are some advantages for making k smaller? What are some advantages for making k larger?

## Practice

### Class Exercises

Examine the following code that fits a knn model using the pokemon data set with $k$ set to $9$. For this example, we are using the full pokemon data set (with all Types), so, we might expect our classification rate to be a bit lower.

```{r}
#| output: false
library(tidyverse)

pokemon <- read_csv(here::here("data/pokemon_full.csv")) |>
  mutate(Type = as.factor(Type))
set.seed(1119)

## scale the quantitative predictors
pokemon_scaled <- pokemon |>
  mutate(across(where(is.numeric), ~ (.x - min(.x)) /
                  (max(.x) - min(.x))))

train_sample <- pokemon_scaled |>
  slice_sample(n = 550)
test_sample <- anti_join(pokemon_scaled, train_sample)

library(kknn)

knn_mod <- kknn(Type ~ HP + Attack + Defense + Speed + SpAtk + SpDef + height + weight,
                train = train_sample, test = test_sample, k = 9)
knn_mod

tab <- table(knn_mod$fitted.values, test_sample$Type)
sum(diag(tab)) / sum(tab)
```

If we want to automate generating a classification rate for a knn model with particular predictors, we have two major choices: write a function and `map()` across different values of $k$ or loop through different values of $k$ with a `for` loop.

__Class Exercise 1__. First, we will take a functional programming approach, by writing a function and "mapping" different values of $k$ through that function. The following code writes a function called `get_class_rate` that has just a single argument: `k_val`

```{r}
#| echo: true
#| output: false

get_class_rate <- function(k_val) {
  knn_mod <- kknn(Type ~ HP + Attack + Defense + Speed + SpAtk + SpDef + height + weight,
                  train = train_sample, test = test_sample, k = k_val)
  
  tab <- table(knn_mod$fitted.values, test_sample$Type)
  class_rate <- sum(diag(tab)) / sum(tab)
  
  return(class_rate)
}
```

a. Run the code and then test the function by running `get_class_rate(k_val = 10)`, which should return the classification rate using `10` nearest neighbors.

b. Together, we will define a vector of k values that we want to map through the function and then write code to perform the mapping using the `map()` function from the `purrr` package.

```{r}
#| echo: false
#| output: false
k_vec <- 1:100
map(k_vec, get_class_rate) |> unlist()
```

c. Put the classification rates, along with the vector of k values into a `tibble()`.

```{r}
#| echo: false
#| output: false
class_rates <- map(k_vec, get_class_rate) |> unlist()
class_df <- tibble(k_values = k_vec,
                   class_rates = class_rates)
```

d. Make a line plot that shows how the classification rate changes for different values of `k`.

```{r}
#| echo: false
#| output: false
ggplot(data = class_df, aes(x = k_values, y = class_rates)) +
  geom_line() +
  theme_minimal()
```

The code below gives an equivalent way to map or loop through values of $k$ using a `for` loop. If you have taken CS 140, you should be able to see a lot of similarities between how loops are defiend in `R` and how they are defined in Python.

```{r}
#| output: false

## define an empty vector to store results
class_rate <- double()

## define values of k that we want to loop through
k_vec <- 1:100

for (i in 1:100) {
  knn_mod <- kknn(Type ~ HP + Attack + Defense + Speed + SpAtk + SpDef + height + weight,
                  train = train_sample, test = test_sample, k = k_vec[i])
    
  knn_mod
  
  tab <- table(knn_mod$fitted.values, test_sample$Type)
  
  ## for the ith value of k_vec, store the classification rate as 
  ## the ith value of class_rate
  class_rate[i] <- sum(diag(tab)) / sum(tab)
}

class_rate
```

### Your Turn

There will be no your turn exercises for this section. Instead, you will apply some of these concepts in your third project.






